---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# taxreturn

<!-- badges: start -->
[![Travis build status](https://travis-ci.org/alexpiper/taxreturn.svg?branch=master)](https://travis-ci.org/alexpiper/taxreturn)
<!-- badges: end -->

taxreturn is an R package for fetching DNA barcode sequences and associated taxonomic annotations from public databases such as the Barcode of Life Database (BOLD) and NCBI GenBank, curating these sequences and formatting them into training sets compatible with popular taxonomic classifiers used for metabarcoding and marker gene analysis.

## Installation

This package is still in development and not yet available on CRAN.  You can install development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("alexpiper/taxreturn")
```
## Examples

This is a basic example which shows you how to solve a common problem:

```{r example}
library(taxreturn)
## basic example code
```
